---
title: "Synthetic data generation for income, education, and environment"
output: html_document
---

## Goal

This document generates **200 observations** for three variables:

1. **income**  
   - positively skewed  
   - includes a few outliers  
   - meant to resemble an income-like distribution

2. **education**  
   - has a **similar skewed shape** to income  
   - measured on a **different scale**

3. **environment**  
   - approximately normal  
   - constructed so that its correlation with **income** is **exactly -0.10** in the sample

The code also verifies the final distributions and correlations.

```{r setup, message=FALSE, warning=FALSE}
set.seed(1234)
n <- 27
rho <- -0.10
```

# Step 1: Create a skewed variable for income

Income is generated from a log-normal distribution, which is commonly used for right-skewed variables.
To create a few outliers, two randomly chosen observations are multiplied by larger factors.

```{r}
# Base skewed income distribution
income <- rlnorm(n, meanlog = 10, sdlog = 0.45)

# Add a few outliers
outlier_id <- sample(1:n, 5)
income[outlier_id] <- income[outlier_id] * c(2.8, 3.5)

summary(income)
```

# Step 2: Create education with a similar shape but a different scale

To make education look similar to income, start from the same latent skewed pattern, then rescale it and add a little noise.
This preserves the broad right-skewed shape while placing the variable on a different magnitude.

```{r}
# Create a similar skewed variable on a different scale
education <- 8 + 0.00035 * income + rnorm(n, mean = 0, sd = 1.2)

# Keep the scale realistic and positive
education <- pmax(education, 1)

# change few observations to reduce correlation with income
randompick <- sample(1:n, 8)
education[randompick] <- education[randompick] + c(sample(-20:20, 5, replace = TRUE))

summary(education)
```

# Step 3: Create environment as a normal variable with correlation -0.10 with income

To achieve a sample correlation of exactly -0.10, the following method is used:

standardize income
generate an independent random normal vector
remove any sample correlation between that random vector and income
combine them using the target correlation formula

This ensures the resulting standardized variable has the desired sample correlation with income.

```{r}
# Standardize income
x <- as.numeric(scale(income))

# Independent normal noise
z <- rnorm(n)

# Remove sample correlation between z and x
z_orth <- resid(lm(z ~ x))

# Standardize orthogonal component
z_orth <- as.numeric(scale(z_orth))

# Construct standardized environment with exact sample correlation rho with x
environment_std <- rho * x + sqrt(1 - rho^2) * z_orth

# Put on a convenient scale
environment <- 50 + 10 * environment_std

summary(environment)
```

# Step 4: Combine into a data frame

```{r}
dat <- data.frame(
  income = round(income, 2),
  education = round(education, 2),
  environment = round(environment, 2)
)
```

# Step 5: Check the correlation matrix

The key requirement is that income and environment have a correlation of -0.10.

```{r, echio=FALSE}
cor(dat)
```

# Step 6: Check distributions visually

These plots help confirm that:

income is right-skewed with a few outliers
education has a similar skewed shape on a different scale
environment looks approximately normal

```{r, fig.height=4, fig.width=12, echo=FALSE}
par(mfrow=c(1,3))
hist(dat$income, main="Income", xlab="Income", col="lightblue", border="white")
hist(dat$education, main="Education", xlab="Education", col="lightgreen", border="white")
hist(dat$environment, main="Environment", xlab="Environment", col="lightcoral", border="white")
```

# Step 7: Export the data

Finally, we can save the generated dataset to a CSV file for future use.    
```{r}
write.csv(dat, "synthetic_data.csv", row.names = FALSE)
```
