---
title: "AMPI: Normalization and Aggregation"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(reshape2)
```

## 1. Data Loading and Exploration

We begin by loading the synthetic dataset and inspecting its structure through correlation analysis and summary statistics.

```{r load-data, echo=FALSE}
data <- read.csv("synthetic_data.csv")

cor(data)
summary(data)
```

The histograms below show the raw distribution of each indicator before any transformation is applied.

```{r, fig.height=4, fig.width=12, echo=FALSE}
par(mfrow=c(1,3))
hist(data$income, main="Income", xlab="Income", col="lightblue", border="white")
hist(data$education, main="Education", xlab="Education", col="lightgreen", border="white")
hist(data$environment, main="Environment", xlab="Environment", col="lightcoral", border="white")
```

### 1.1 Scope of the Analysis
The dataset includes three indicators — **income**, **education**, and **environment** — measured for a set of countries. The goal is to construct a composite index that aggregates these dimensions into a single score reflecting overall performance. The index must be positive (i.e. multidimensional development) and we assume that all 3 indicators are measured in a way that higher values indicate better performance (e.g., higher income, better education outcomes, cleaner environment).

---

## 2. Normalization: Constrained Min-Max (AMPI Method)

### 2.1 Rationale

Raw indicators are measured on different scales and units, making direct comparison or aggregation meaningless. We therefore normalize each indicator onto a common scale before aggregating.

### 2.2 Goalpost and Range

The normalization uses a **constrained min-max** approach where:

- The **goalpost** (reference point) is the **cross-country median** of each indicator — i.e., all countries aim at reaching the median European value.
- The normalized score is bounded between **70** (worst performer) and **130** (best performer), yielding a symmetric range of ±30 around the central value of 100.

### Formula

For each indicator $x$, the normalized score $z$ is:

$$z_i = 70 + 60 \cdot \frac{x_i - \tilde{x}}{x_{\max} - x_{\min}}$$

Where:

| Symbol | Meaning |
|---|---|
| $x_i$ | Raw value for country $i$ |
| $\tilde{x}$ | Median of the indicator across all countries |
| $x_{\max}$ | Maximum observed value |
| $x_{\min}$ | Minimum observed value |
| $70$ | Floor score (worst performer) |
| $130$ | Ceiling score (best performer) |

The multiplier $60$ reflects the total range $[70, 130]$, so that deviations from the median are rescaled proportionally within that interval.

```{r normalize}
AMPI_norm <- function(x) {
  70 + (60 * ((x - median(x, na.rm = TRUE)) /
    (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))))
}

dat_std <- as.data.frame(apply(data, 2, AMPI_norm))

cor(dat_std)
summary(dat_std)
```

The histograms below show the distribution of each indicator after normalization. All variables now share the same $[70, 130]$ scale.

```{r plot-normalized, echo=FALSE}
dat_long_std <- reshape2::melt(dat_std)
ggplot(dat_long_std, aes(x = value, fill = variable)) +
  geom_histogram(alpha = 0.4, position = "identity", bins = 30) +
  labs(title = "Overlapping Histograms (Normalized Data)",
       x = "Value", y = "Count") +
  theme_minimal()
```

---

## 3. Coefficient of Variation: Before and After Normalization

The **Coefficient of Variation (CV)** measures relative dispersion as the ratio of the standard deviation to the mean:

$$CV = \frac{\sigma}{\bar{x}}$$

Comparing CVs before and after normalization helps verify that the transformation preserves the relative variability structure across indicators.

```{r cv-comparison, echo=FALSE}
cat("CV before normalization:\n")
sapply(data, function(x) sd(x, na.rm = TRUE) / 
    mean(x, na.rm = TRUE))

cat("\nCV after normalization:\n")
sapply(dat_std, function(x) sd(x, na.rm = TRUE) / 
    mean(x, na.rm = TRUE))
```

---

## 4. Aggregation: The AMPI Index

### Formula

The **AMPI (Adjusted Mean Performance Index)** aggregates the normalized indicators for each country using:

$$\text{AMPI}_i = \bar{z}_i - \sigma_i \cdot CV_i$$

Where all quantities are computed **row-wise** (i.e., across indicators for each country $i$):

| Symbol | Meaning |
|---|---|
| $\bar{z}_i$ | Mean of normalized scores across indicators |
| $\sigma_i$ | Standard deviation of normalized scores across indicators |
| $CV_i$ | Coefficient of variation of normalized scores across indicators |

### Interpretation

The additive correction term $\sigma_i \cdot CV_i$ **rewards countries that perform consistently well** across all indicators. A country with high average scores *and* low relative dispersion will receive a smaller penalty than one with an equivalent mean but highly uneven performance across dimensions.

```{r aggregation}
dat_std$AMPI <- rowMeans(dat_std) -
  (apply(dat_std, 1, sd) *
     apply(dat_std, 1, function(x) sd(x, na.rm = TRUE) / 
     mean(x, na.rm = TRUE)))

summary(dat_std$AMPI)
cor(dat_std)
```


---

## 5. Comparison Method: Geometric Mean Aggregation

To benchmark the AMPI approach, we apply a standard **min-max normalization** followed by **geometric mean aggregation** — a common composite index methodology (e.g., used in the UNDP Human Development Index since 2010).


### 5.2 Aggregation: Geometric Mean

#### Formula

The geometric mean aggregates the $k$ normalized indicators for each country $i$ as:

$$GM_i = \left(\prod_{j=1}^{k} z_{ij}\right)^{1/k} = \sqrt[k]{z_{i1} \cdot z_{i2} \cdot \ldots \cdot z_{ik}}$$

#### Why the Geometric Mean?

Compared to an arithmetic mean, the geometric mean **penalizes imbalance** across dimensions: a country with a very low score on any single indicator cannot fully compensate by excelling in others. This property — called **imperfect substitutability** — is often desirable in multidimensional welfare indices.

> ⚠️ **Note:** Because min-max scores can equal exactly **0**, and $\log(0)$ is undefined, a small constant $\varepsilon$ is added before taking the geometric mean to avoid undefined results:
> $$GM_i = \left(\prod_{j=1}^{k} (z_{ij} + \varepsilon)\right)^{1/k}$$

```{r geometric-mean-aggregation}
epsilon <- 1e-6  # small constant to avoid log(0)

dat_std$GM <- apply(dat_std[,c(1:3)], 1, function(x) {
  exp(mean(log(x + epsilon)))
})

dat_std$AM <- rowMeans(dat_std[,c(1:3)])

summary(dat_std$GM)
cor(dat_std)
```

---

## 6. Head-to-Head Comparison: AMPI vs. Geometric Mean

### 6.1 Correlation Between the Two Indices

A high correlation between AMPI and GM would suggest the two methods rank countries similarly despite their different normalization and aggregation choices.

```{r index-correlation, echo=FALSE}
comparison <- data.frame(
  AMPI = dat_std$AMPI,
  GM   = dat_std$GM,
  AM   = dat_std$AM
)

cor(comparison)

ggplot(comparison, aes(x = AMPI, y = GM)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  labs(title = "AMPI vs. Geometric Mean Index",
       x = "AMPI Score", y = "Geometric Mean Score") +
  theme_minimal()
```

### 6.2 Regression $R^2$ Comparison Table

We compare how strongly each indicator predicts the composite index under each method:
```{r regressions-gm, echo=FALSE}
```{r regressions}
modX1 <- lm(AMPI ~ income,      data = dat_std)
modX2 <- lm(AMPI ~ education,   data = dat_std)
modX3 <- lm(AMPI ~ environment, data = dat_std)

modGM_X1 <- lm(GM ~ income,      data = dat_std)
modGM_X2 <- lm(GM ~ education,   data = dat_std)
modGM_X3 <- lm(GM ~ environment, data = dat_std)

modAM_X1 <- lm(AM ~ income,      data = dat_std)
modAM_X2 <- lm(AM ~ education,   data = dat_std)
modAM_X3 <- lm(AM ~ environment, data = dat_std)
```

```{r r2-comparison-table, echo=FALSE}
r2 <- function(model) summary(model)$r.squared

r2_table <- data.frame(
  Indicator = c("Income", "Education", "Environment"),
  AMPI_R2   = round(c(r2(modX1), r2(modX2), r2(modX3)), 3),
  GM_R2     = round(c(r2(modGM_X1), r2(modGM_X2), r2(modGM_X3)), 3),
  AM_R2     = round(c(r2(modAM_X1), r2(modAM_X2), r2(modAM_X3)), 3)
)

knitr::kable(r2_table,
             col.names = c("Indicator", "AMPI $R^2$", "GM $R^2$", "AM $R^2$"),
             caption = "Comparison of indicator-level $R^2$ values across aggregation methods")
```

### 6.3 Summary Comparison Table

```{r summary-table, echo=FALSE}
knitr::kable(
  data.frame(
    Feature             = c("Normalization range", "Central goalpost",
                            "Aggregation method", "Penalizes imbalance",
                            "Sensitive to outliers", "Undefined for zero values"),
    AMPI                = c("[70, 130]", "Mean", 
                            "Arithmetic mean + SD × CV", "Yes (via CV term)",
                            "Moderate", "No"),
    Geometric_Mean      = c("[0, 1]", "None (observed extremes)",
                            "Geometric mean", "Yes (multiplicative)",
                            "High", "Yes (requires ε adjustment)")
  ),
  col.names = c("Feature", "AMPI", "Min-Max + Geometric Mean"),
  caption   = "Methodological comparison of the two composite index approaches"
)
```

---

## 8. Correspondence Table: Raw Indicators, AMPI and Geometric Mean Rankings

We build a unified ranking table that displays, for each country, both the raw indicator values and the rank assigned by each composite index. This allows a direct visual audit of how normalization and aggregation choices affect country standings.

```{r ranking-table, echo=FALSE}
# Assemble raw indicators + both indices into one dataframe
ranking_df <- data.frame(
  Country     = rownames(data),  # adjust if your country identifier is a column
  Income      = data$income,
  Education   = data$education,
  Environment = data$environment,
  AMPI        = dat_std$AMPI,
  GM          = dat_std$GM
)

# Compute rankings (rank 1 = best performer)
ranking_df$Rank_Income      <- rank(-ranking_df$Income,      ties.method = "min")
ranking_df$Rank_Education   <- rank(-ranking_df$Education,   ties.method = "min")
ranking_df$Rank_Environment <- rank(-ranking_df$Environment, ties.method = "min")
ranking_df$Rank_AMPI        <- rank(-ranking_df$AMPI,        ties.method = "min")
ranking_df$Rank_GM          <- rank(-ranking_df$GM,          ties.method = "min")

```

### 8.1 Rank Disagreement Between AMPI and Geometric Mean

Countries where the two methods disagree most reveal the **sensitivity of rankings to methodological choices**. A large rank gap signals that a country's position is not robust — it depends heavily on how normalization and aggregation are specified.

```{r rank-disagreement, echo=FALSE}
ranking_df$Rank_Gap <- abs(ranking_df$Rank_AMPI - ranking_df$Rank_GM)

disagreement <- ranking_df[order(-ranking_df$Rank_Gap),c(1,7:11)]

knitr::kable(
  head(disagreement,20),
  row.names = FALSE,
  caption   = "Top 20Countries ranked by disagreement between AMPI and Geometric Mean index"
)
```
