---
title: "AMPI: Normalization and Aggregation"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(reshape2)
```

## 1. Data Loading and Exploration

We begin by loading the synthetic dataset and inspecting its structure through correlation analysis and summary statistics.

```{r load-data, echo=FALSE}
data <- read.csv("synthetic_data.csv")

cor(data)
summary(data)
```

The histograms below show the raw distribution of each indicator before any transformation is applied.

```{r, fig.height=4, fig.width=12, echo=FALSE}
par(mfrow=c(1,3))
hist(data$income, main="Income", xlab="Income", col="lightblue", border="white")
hist(data$education, main="Education", xlab="Education", col="lightgreen", border="white")
hist(data$environment, main="Environment", xlab="Environment", col="lightcoral", border="white")
```

### 1.1 Scope of the Analysis
The dataset includes three indicators — **income**, **education**, and **environment** — measured for a set of countries. The goal is to construct a composite index that aggregates these dimensions into a single score reflecting overall performance. The index must be positive (i.e. multidimensional development) and we assume that all 3 indicators are measured in a way that higher values indicate better performance (e.g., higher income, better education outcomes, cleaner environment).

---

## 2. Normalization Comparison

### 2.1 Rationale

Raw indicators are measured on different scales and units, making direct comparison or aggregation meaningless. We therefore normalize each indicator onto a common scale before aggregating.

### 2.2 Standardization
A common approach is **z-score standardization**, which transforms each variable to have a mean of 0 and a standard deviation of 1:
$$z_i = \frac{x_i - \bar{x}}{\sigma}$$

Where:

- $x_i$ is the raw value for country $i$
- $\bar{x}$ is the mean of the indicator across all countries
- $\sigma$ is the standard deviation of the indicator across all countries
This method centers the data around zero and rescales it based on variability, but it does not set a fixed range or reference point for interpretation.

```{r std, echo=FALSE}
dat_std <- as.data.frame(scale(data))

```

### 2.3 Min-Max Normalization
Another common method is **min-max normalization**, which rescales the data to a fixed range, typically [0, 1]:
$$z_i = \frac{x_i - x_{\min}}{x_{\max} - x_{\min}}$$
Where:
- $x_i$ is the raw value for country $i$
- $x_{\min}$ is the minimum observed value of the indicator across all countries
- $x_{\max}$ is the maximum observed value of the indicator across all countries
This method preserves the relative position of each country within the observed range but can be sensitive to outliers and does not set a meaningful reference point for performance.

```{r minmax, echo=FALSE}
minmax <- function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}
dat_mm <- as.data.frame(apply(data, 2, minmax))

```

### 2.4 AMPI Normalization: Constrained Min-Max with Goalpost
The normalization uses a **constrained min-max** approach where:
- The **goalpost** (reference point) is the **cross-country median** of each indicator — i.e., all countries aim at reaching the median European value.
- The normalized score is bounded between **70** (worst performer) and **130** (best performer), yielding a symmetric range of ±30 around the central value of 100.

For each indicator $x$, the normalized score $z$ is:
$$z_i = 70 + 60 \cdot \frac{x_i - \tilde{x}}{x_{\max} - x_{\min}}$$

Where:
- $x_i$ | Raw value for country $i$ 
- $\tilde{x}$ | Median of the indicator across all countries 
- $x_{\max}$ | Maximum observed value 
- $x_{\min}$ | Minimum observed value 
- $70$ | Floor score (worst performer) 
- $130$ | Ceiling score (best performer) 

The multiplier $60$ reflects the total range $[70, 130]$, so that deviations from the median are rescaled proportionally within that interval.

```{r ampinorm, echo=FALSE}
AMPI_norm <- function(x) {
  70 + (60 * ((x - median(x, na.rm = TRUE)) /
    (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))))
}

dat_ampi <- as.data.frame(apply(data, 2, AMPI_norm))
```

### 2.5 Comparison of Normalization Methods
Comparing the distributions of the normalized indicators under each method reveals how the choice of normalization affects the relative positioning of countries and the interpretability of scores. The AMPI normalization centers scores around 100 with a fixed range, while standardization centers around 0 without a fixed range, and min-max normalization rescales to [0, 1] based on observed extremes.

Each method has different implications for how we interpret the scores and how sensitive they are to outliers or the distribution of the raw data. The AMPI method, by using a median goalpost and a fixed range, emphasizes relative performance against a central benchmark while allowing for a clear interpretation of scores as "above" or "below" the median. The standardization method emphasizes deviations from the mean in terms of standard deviations, while the min-max method emphasizes relative position within the observed range.

```{r plot-normalized, echo=FALSE}
# plot using facet in ggplot the overlapped histograms of the three normalized indicators for each methodology
dat_std_melt <- melt(dat_std, variable.name = "Indicator", value.name = "Z_Score")
dat_mm_melt <- melt(dat_mm, variable.name = "Indicator", value.name = "Z_Score")
dat_ampi_melt <- melt(dat_ampi, variable.name = "Indicator", value.name = "Z_Score")
dat_std_melt$Method <- "Standardization"
dat_mm_melt$Method <- "Min-Max"
dat_ampi_melt$Method <- "AMPI Normalization"
dat_all <- rbind(dat_std_melt, dat_mm_melt, dat_ampi_melt)
ggplot(dat_all, aes(x = Z_Score, fill = Indicator)) +
  geom_histogram(alpha = 0.5, position = "identity", bins = 30) +
  facet_wrap(~ Method, scales = "free") +
  labs(title = "Comparison of Normalization Methods",
       x = "Normalized Score", y = "Frequency") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The **Coefficient of Variation (CV)** measures relative dispersion as the ratio of the standard deviation to the mean:
$$CV = \frac{\sigma}{\bar{x}}$$

Comparing CVs before and after normalization helps verify that the transformation preserves the relative variability structure across indicators.

```{r cv-comparison, echo=FALSE}
cv_raw <- apply(data, 2, function(x) sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE))
cv_std <- apply(dat_std, 2, function(x) sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE))
cv_mm <- apply(dat_mm, 2, function(x) sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE))
cv_ampi <- apply(dat_ampi, 2, function(x) sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE))
cv_comparison <- data.frame(
  Indicator = names(cv_raw),
  Raw_CV    = round(cv_raw, 3),
  Std_CV    = round(cv_std, 3),
  MM_CV     = round(cv_mm, 3),
  AMPI_CV   = round(cv_ampi, 3)
)
knitr::kable(cv_comparison,
             col.names = c("Indicator", "Raw CV", "Standardization CV", "Min-Max CV", "AMPI CV"),
             caption = "Comparison of Coefficient of Variation across normalization methods")
```

---

## 3. Comparison of Aggregation Methods

### 3.1 Rationale
After normalization, we need to aggregate the multiple indicators into a single composite index. The choice of aggregation method affects how we interpret the overall score and how sensitive it is to imbalances across dimensions. In the following section we depart from deta normalized via **AMPI Constraied Min-Max** and apply two different aggregation methods.

### 3.2 Additive Aggregation
The **additive aggregation** method computes the composite index as a weighted sum of the normalized indicators:
$$I_i = \sum_{j=1}^{k} w_j z_{ij}$$
Where:
- $I_i$ is the composite index score for country $i$
- $w_j$ is the weight assigned to indicator $j$ (often set to 1/k for equal weighting)
- $z_{ij}$ is the normalized score of country $i$ on indicator $j$
This method assumes **perfect substitutability** between indicators: a low score on one indicator can be fully compensated by a high score on another. It is straightforward and easy to interpret, but it may not capture the multidimensional nature of development if we want to penalize imbalances across dimensions.

```{r sum, echo=FALSE}
dat_final <- cbind(data, dat_ampi)

dat_final$Sum <- rowSums(dat_final[,c(4:6)])

```

### 3.3 Power-Mean Aggregation
The **power-mean aggregation** method generalizes the arithmetic mean by introducing a parameter $p$ that controls the degree of substitutability:
$$I_i = \left(\frac{1}{k} \sum_{j=1}^{k} z_{ij}^p\right)^{1/p}$$
Where:
- $I_i$ is the composite index score for country $i$
- $p$ is the power parameter that determines the aggregation behavior
- $z_{ij}$ is the normalized score of country $i$ on indicator $j$
When $p = 1$, this reduces to the arithmetic mean (perfect substitutability). When $p < 1$, it penalizes imbalance more strongly, and when $p > 1$, it allows for more compensation between indicators. The choice of $p$ reflects normative judgments about how we value balance across dimensions.
```{r power-mean, echo=FALSE}
# we loop to create a composite indicator aggregated via power mean for different values of p, we can then compare the results with the additive aggregation method
# We compare the aritmetic mean, the geometric mean, the armonic mean
p_mean <- function(x,p){
  
}



```


## 5. Comparison Method: Geometric Mean Aggregation

To benchmark the AMPI approach, we apply a standard **min-max normalization** followed by **geometric mean aggregation** — a common composite index methodology (e.g., used in the UNDP Human Development Index since 2010).


### 5.2 Aggregation: Geometric Mean

#### Formula

The geometric mean aggregates the $k$ normalized indicators for each country $i$ as:

$$GM_i = \left(\prod_{j=1}^{k} z_{ij}\right)^{1/k} = \sqrt[k]{z_{i1} \cdot z_{i2} \cdot \ldots \cdot z_{ik}}$$

#### Why the Geometric Mean?

Compared to an arithmetic mean, the geometric mean **penalizes imbalance** across dimensions: a country with a very low score on any single indicator cannot fully compensate by excelling in others. This property — called **imperfect substitutability** — is often desirable in multidimensional welfare indices.

> ⚠️ **Note:** Because min-max scores can equal exactly **0**, and $\log(0)$ is undefined, a small constant $\varepsilon$ is added before taking the geometric mean to avoid undefined results:
> $$GM_i = \left(\prod_{j=1}^{k} (z_{ij} + \varepsilon)\right)^{1/k}$$

```{r geometric-mean-aggregation}
epsilon <- 1e-6  # small constant to avoid log(0)

dat_std$GM <- apply(dat_std[,c(1:3)], 1, function(x) {
  exp(mean(log(x + epsilon)))
})

dat_std$AM <- rowMeans(dat_std[,c(1:3)])

summary(dat_std$GM)
cor(dat_std)
```

---

## 6. Head-to-Head Comparison: AMPI vs. Geometric Mean

### 6.1 Correlation Between the Two Indices

A high correlation between AMPI and GM would suggest the two methods rank countries similarly despite their different normalization and aggregation choices.

```{r index-correlation, echo=FALSE}
comparison <- data.frame(
  AMPI = dat_std$AMPI,
  GM   = dat_std$GM,
  AM   = dat_std$AM
)

cor(comparison)

ggplot(comparison, aes(x = AMPI, y = GM)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  labs(title = "AMPI vs. Geometric Mean Index",
       x = "AMPI Score", y = "Geometric Mean Score") +
  theme_minimal()
```

### 6.2 Regression $R^2$ Comparison Table

We compare how strongly each indicator predicts the composite index under each method:
```{r regressions-gm, echo=FALSE}
```{r regressions}
modX1 <- lm(AMPI ~ income,      data = dat_std)
modX2 <- lm(AMPI ~ education,   data = dat_std)
modX3 <- lm(AMPI ~ environment, data = dat_std)

modGM_X1 <- lm(GM ~ income,      data = dat_std)
modGM_X2 <- lm(GM ~ education,   data = dat_std)
modGM_X3 <- lm(GM ~ environment, data = dat_std)

modAM_X1 <- lm(AM ~ income,      data = dat_std)
modAM_X2 <- lm(AM ~ education,   data = dat_std)
modAM_X3 <- lm(AM ~ environment, data = dat_std)
```

```{r r2-comparison-table, echo=FALSE}
r2 <- function(model) summary(model)$r.squared

r2_table <- data.frame(
  Indicator = c("Income", "Education", "Environment"),
  AMPI_R2   = round(c(r2(modX1), r2(modX2), r2(modX3)), 3),
  GM_R2     = round(c(r2(modGM_X1), r2(modGM_X2), r2(modGM_X3)), 3),
  AM_R2     = round(c(r2(modAM_X1), r2(modAM_X2), r2(modAM_X3)), 3)
)

knitr::kable(r2_table,
             col.names = c("Indicator", "AMPI $R^2$", "GM $R^2$", "AM $R^2$"),
             caption = "Comparison of indicator-level $R^2$ values across aggregation methods")
```

### 6.3 Summary Comparison Table

```{r summary-table, echo=FALSE}
knitr::kable(
  data.frame(
    Feature             = c("Normalization range", "Central goalpost",
                            "Aggregation method", "Penalizes imbalance",
                            "Sensitive to outliers", "Undefined for zero values"),
    AMPI                = c("[70, 130]", "Mean", 
                            "Arithmetic mean + SD × CV", "Yes (via CV term)",
                            "Moderate", "No"),
    Geometric_Mean      = c("[0, 1]", "None (observed extremes)",
                            "Geometric mean", "Yes (multiplicative)",
                            "High", "Yes (requires ε adjustment)")
  ),
  col.names = c("Feature", "AMPI", "Min-Max + Geometric Mean"),
  caption   = "Methodological comparison of the two composite index approaches"
)
```

---

## 8. Correspondence Table: Raw Indicators, AMPI and Geometric Mean Rankings

We build a unified ranking table that displays, for each country, both the raw indicator values and the rank assigned by each composite index. This allows a direct visual audit of how normalization and aggregation choices affect country standings.

```{r ranking-table, echo=FALSE}
# Assemble raw indicators + both indices into one dataframe
ranking_df <- data.frame(
  Country     = rownames(data),  # adjust if your country identifier is a column
  Income      = data$income,
  Education   = data$education,
  Environment = data$environment,
  AMPI        = dat_std$AMPI,
  GM          = dat_std$GM
)

# Compute rankings (rank 1 = best performer)
ranking_df$Rank_Income      <- rank(-ranking_df$Income,      ties.method = "min")
ranking_df$Rank_Education   <- rank(-ranking_df$Education,   ties.method = "min")
ranking_df$Rank_Environment <- rank(-ranking_df$Environment, ties.method = "min")
ranking_df$Rank_AMPI        <- rank(-ranking_df$AMPI,        ties.method = "min")
ranking_df$Rank_GM          <- rank(-ranking_df$GM,          ties.method = "min")

```

### 8.1 Rank Disagreement Between AMPI and Geometric Mean

Countries where the two methods disagree most reveal the **sensitivity of rankings to methodological choices**. A large rank gap signals that a country's position is not robust — it depends heavily on how normalization and aggregation are specified.

```{r rank-disagreement, echo=FALSE}
ranking_df$Rank_Gap <- abs(ranking_df$Rank_AMPI - ranking_df$Rank_GM)

disagreement <- ranking_df[order(-ranking_df$Rank_Gap),c(1,7:11)]

knitr::kable(
  head(disagreement,20),
  row.names = FALSE,
  caption   = "Top 20Countries ranked by disagreement between AMPI and Geometric Mean index"
)
```
